---
title       : R and statistics
description : Basics of R, the amazing statistical programming language. Do not be afraid of the art of programming!

--- type:NormalExercise lang:r xp:50 skills:1 
## What is R?

R is a programming language designed for statistical analysis. R is completely open source and all analysis done with R are reproduciple and easy to share. R is flexible, powerful and includes implementations of the latest research. 

Because of these reasons, R is a very popular language among the scientific community. R is however also widely [used](http://blog.revolutionanalytics.com/2014/05/companies-using-r-in-2014.html) for data analysis and visualisations by companies such as Google, Facebook, Twitter and Reaktor. R is perfect for data science.

Researchers and experts often use programming to visualize and explore data and implement new statistical methods. The purpose of this course is to expose you to the basics of programming and data analysis with R.

Follow the instructions below to complete the exercise.

*** =instructions
- Take a peek at the code written on the right in script.R. When you opened this exercise, the code gave R the instructions to (1) load a dataset and (2) produce the graphic next to it.
- Could you tell which lines of code relates to (1) or (2)? Hint: A hashtag marks a line as a comment line.
- When you're done pondering, simply click 'Submit Answer' and move on to the first exercise!

***= hint
- Simply press 'Submit Answer' when you're ready to move forward.


*** =pre_exercise_code
```{r}
data_url <- "http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS-data.txt"

students2014 <- read.table(data_url, sep="\t", header=TRUE)

with(students2014, plot(x = kenka, y = pituus, col = sukup, pch = 20, main = "Scatter plot of student shoesize and height \n(color indicates sex)"))
```

*** =sample_code
```{r}
# This is an example of the power of R

# Create a character object holding data location
data_url <- "http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS-data.txt"

# Load and save data with the read.table() function
students2014 <- read.table(data_url, sep = "\t", header = TRUE)

# Draw a graph of student shoesize and hight using plot()
with(students2014, plot(x = kenka, y = pituus, col = sukup, pch = 20, main = "Scatter plot of student shoesize and height \n(color indicates sex)"))


```

*** =solution
```{r}
# This is an example of the power of R

# Create a character object holding data location
data_url <- "http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS-data.txt"

# Load and save data with the read.table() function
students2014 <- read.table(data_url, sep = "\t", header = TRUE)

# Draw a graph of student shoesize and hight using plot()
with(students2014, plot(x = kenka, y = pituus, col = sukup, pch = 20, main = "Scatter plot of student shoesize and height \n(color indicates sex)"))

```

*** =sct
```{r}
success_msg("Great! Move on to the first exercise.")
```

--- type:NormalExercise lang:r xp:100 skills:1
## Basic tools

On your right you see the R editor area - the script - and below that the R console. The editor area is just a simple text editor where you write code - just like text.

You can first write code to the editor area and then tell R that you want to execute a line of code where your cursor currently is by pressing `Ctrl + Enter` (`Cmd + Enter` on a mac). Input and output will then appear in the console.

It is also possible to write code directly to the console and use `Enter` to execute, but working with the script area is preferred.

Try it!
  
*** =instructions
- Type "Hello world!" in the editor area. Use quotation marks.
- Press `Ctrl + Enter` to execute your "Hello world!" - code
- Make sure to use a capital "H" and an exclamation mark.
- Click 'Submit Answer' when done.

*** =hint
- Remember to use quotation marks and write the sentence exactly as instructed.


*** =pre_exercise_code
```{r}
# no pec
```


*** =sample_code
```{r}
# This is the R editor!

# A hashtag at the beginning of the line defines the line as a comment

# Write your code below





# Below is the R console, where you will see output
```

*** =solution
```{r}
# This is the R editor!

# A hashtag at the beginning of the line defines the line as a comment

# write your code below
"Hello world!"




# Below is the R console, where you will see output
```

*** =sct
```{r}
test_student_typed("Hello world!", not_typed_msg="Please type 'Hello world!' in the editor.")

test_error()
success_msg("Great work! You have now executed your first line of code :)")
```


--- type:NormalExercise lang:r xp:100 skills:1
## Arithmetics

Let's continue with something simple. R can do amazing things like scrape websites and draw beautiful graphics but it can also do basic calculations. Consider the following arithmetic operators:

- Addition: `+`
- Subtraction: `-`
- Multiplication: `*`
- Division: `/`
- Exponentiation: `^`

The `^` operator raises the number to its left to the power of the number to its right: for example `3^2` is 9. With this knowledge, follow the instructions below to complete the exercise.

*** =instructions
- Study and execute the examples in the R script. (Use `Ctrl + Enter` or `Cmd + Enter`)
- Type `2^5` in the editor and calculate 2 to the power 5.
- When done, click 'Submit Answer'.

*** =hint
- If you have trouble accessing the `^` -sign, you can also use double star `**`.


*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# An addition
5 + 5 

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5) / 2 

# Exponentiation



```

*** =solution
```{r}
# An addition
5 + 5

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5) / 2 

# Exponentiation
2 ^ 5

```

*** =sct
```{r}
test_output_contains("2^5", incorrect_msg = "The exponentiation example is not correct. Please write `2 ^ 5` on a new line.")

success_msg("Great work! Head over to the next exercise.")
```

--- type:NormalExercise lang:r xp:100 skills:1 
## Objects

Here's where things start to get interesting. In R you can create and operate on things called *objects*. 
An object is something that can store information such as numerical values and names. Once an object is created, it will be stored in memory and the information it contains will be available to you later.

Objects are created using the assign operator: `<-` (< and -). The value of an object can be printed by typing it's name.


*** =instructions
- Execute the lines that create and operate on `my_character_object` and `my_numeric_object`.
- Assign your name or nickname to `my_name`. Use quotation marks.
- Assign a positive message to yourself to `my_message`. Use quotation marks.
- Click 'Submit Answer'.

*** =hint
- Remember that `Ctrl + Enter` executes a single row.

*** =pre_exercise_code
```{r}
# no pec
```


*** =sample_code
```{r}

# Create an object
my_character_object <- "Hi there!"

# Print the contents of the object
my_character_object

# Create another object
my_numeric_object <- 5 + 5

# Do further calculations with the object
my_numeric_object / 5

# Override the value of an object by assigning a new value to it
my_character_object <- my_numeric_object
my_character_object

# Create character objects my_name and my_message here. Use quotation marks.
my_name <- 
my_message <- 

```

*** =solution
```{r}

# Create an object
my_character_object <- "Hi there!"

# Print the contents of the object
my_character_object

# Create another object
my_numeric_object <- 5 + 5

# Do further calculations with the object
my_numeric_object / 5

# Override the value of an object by assigning a new value to it
my_character_object <- my_numeric_object
my_character_object

# Create character objects my_name and my_message here. Use quotation marks.
my_name <- "Hey You"
my_message <- "You are awsome!"

```

*** =sct
```{r}
test_object("my_name", eval = F, incorrect_msg = "Did you assign your name to `my_name`?")
test_object("my_message", eval = F, incorrect_msg = "Did you assign a message to `my_message`?")

test_error(incorrect_msg = "Did you use quotation marks to assign your name and a message to the objects `my_name` and `my_message`")
success_msg(paste0("Great job ", my_name, ". ", my_message, "!"))
```


--- type:NormalExercise lang:r xp:100 skills:1
## Functions

In R you operate on objects using *functions*. A function is a special kind of object that can be *called* by typing it's name and giving it the necessary *arguments* inside parenthesis. When a function is called, it will perform an action.

R has many functions ready for you to use. Functions are how you're actually going to do all the magic.

Here we will do a little bit of magic using data that will become more familiar to you later on.  

*** =instructions
- Study and execute the examples in the R script. Don't worry about the `$` -sign for now, we will get to that later.
- Follow the examples and compute the mean age of students.
- When done, click 'Submit Answer'.

*** =hint
- Use the function `mean()` and the object `student_age` like in the example on line 7 of script.R.


*** =pre_exercise_code
```{r}
students2014 <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS-data.txt", sep="\t", header=TRUE)

# keep a couple background variables
students2014 <- students2014[,c("sukup","toita","ika","pituus","kenka","kone")]

# recode kone -variable NA values as factor levels
students2014$kone <- addNA(students2014$kone)

# choose rows without missing values
students2014 <- students2014[complete.cases(students2014),]

# integers to numeric
students2014$ika <- as.numeric(students2014$ika)
students2014$pituus <- as.numeric(students2014$pituus)
students2014$kenka <- as.numeric(students2014$kenka)

```

*** =sample_code
```{r}
# students2014 data is available

# Create object student_height
student_height <- students2014$pituus

# Compute the average height of the students
mean(student_height)

# Create object student_age
student_age <- students2014$ika

# Compute the average age of the students




```

*** =solution
```{r}
# students2014 data is available

# Create object student_height
student_height <- students2014$pituus

# Compute the average height of the students
mean(student_height)

# Create object student_age
student_age <- students2014$ika

# Compute the average age of the students
mean(student_age)



```

*** =sct
```{r}
test_output_contains("mean(student_age)", incorrect_msg = "Please use the `mean()` function to compute the mean of student_age.")

test_error()
success_msg("Excellent! You are making great progress.")
```


--- type:NormalExercise lang:r xp:100 skills:1
## Good arguments

During this course you will operate on many of R's ready-made functions. As we already saw, functions take objects as their arguments and then perform actions using the objects. Using a function is called *calling* it.

Functions usually have more than one possible argument. Some of them can have default values and others need to be specified. The arguments have names, which can (and often should) be used when specifying their values. The recommended style is to specify all but the first argument by name, when calling a function.

*** =instructions
- Execute the example codes where the function `head()` is used to explore the first couple observations in the `students2014` dataset.
- Create object `first_ten_students` by calling `head()` with `n = 10`.

*** =hint
- Remember to separate the arguments with a comma
- The answer is of the form: `first_ten_students <- head(arguments_here)`


*** =pre_exercise_code
```{r}
# load data from web
students2014 <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS-data.txt", sep="\t", header=TRUE)
# keep a couple background variables
students2014 <- students2014[,c("sukup","toita","ika","pituus","kenka","kone")]
# recode kone variables missing values as factor levels
students2014$kone <- addNA(students2014$kone)
# keep only rows without missing values
students2014 <- students2014[complete.cases(students2014),]
# integers to numeric
students2014$ika <- as.numeric(students2014$ika)
students2014$pituus <- as.numeric(students2014$pituus)
students2014$kenka <- as.numeric(students2014$kenka)

```


*** =sample_code
```{r}
# students2014 is available

# Use the function head() on the object students2014
head(students2014, n = 4)

# Argument 'n' of head() has a default value 6
head(students2014)

# These calls are identical. Try them.
head(students2014, n = 3) # recommended style
head(students2014, 3)     # style usually not recommended
head(n = 3, students2014) # style not recommended

# Save the first ten observations of the data
first_ten_students <- 


```

*** =solution
```{r}
# students2014 is available

# Use the function head() on the object students2014
head(students2014, n = 4)

# Argument 'n' of head() has a default value 6
head(students2014)

# These calls are identical. Try them.
head(students2014, n = 3) # recommended style
head(students2014, 3)     # style usually not recommended
head(n = 3, students2014) # style not recommended

first_ten_students <- head(students2014, n = 10)


```

*** =sct
```{r}
test_error(incorrect_msg = "Your code seems to produce an error. Did you create the object `first_ten_students` as instructed?")

test_object("first_ten_students")

success_msg("Great job!")
```


--- type:NormalExercise lang:r xp:100 skills:1
## Data frames

In R, the basic idea is to use *functions* to operate on objects. The most common object to operate on is the `data frame` object, which stores data in tabular form.  

The columns in data frames are vectors. Data frames can contain multiple vectors with different data types. In this chapter you will learn more about vectors and data types.

Here you will get to apply some functions that are helpful for inspecting data: `head()` and `str()`. You will also learn how to access the columns of a data frame using the $-sign.  

*** =instructions
- Execute the function `head()` and see the results
- Inspect the data types in the `students2014` data with the function `str()`
- Use the $-sign to access the students shoe size (`kenka`) in `students2014` data
- Note that `students2014$ika` prints out all the ages of the students in `students2014`

*** =hint
- The data is stored in object `students2014`

*** =pre_exercise_code
```{r}
students2014 <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS-data.txt", sep="\t", header=TRUE)

# keep a couple background variables
students2014 <- students2014[,c("sukup","toita","ika","pituus","kenka","kone")]

# recode kone -variable NA values as factor levels
students2014$kone <- addNA(students2014$kone)

# choose rows without missing values
students2014 <- students2014[complete.cases(students2014),]

# integers to numeric
students2014$ika <- as.numeric(students2014$ika)
students2014$pituus <- as.numeric(students2014$pituus)
students2014$kenka <- as.numeric(students2014$kenka)
```

*** =sample_code
```{r}
# students2014 is available

# Let's view the first 6 students from the course participants data
head(students2014)

# View the structure of students2014
str(students2014)

# You can access the variables in the students2014 data by using the $-sign
students2014$ika

# Access the shoe sizes of the students


```

*** =solution
```{r}
# students2014 is available

# Let's view the first 6 students from the course participants data
head(students2014)

# View the structure of students2014
str(students2014)

# You can access the variables in the students2014 data by using the $-sign
students2014$ika

# Access the shoe sizes of the students
students2014$kenka
```

*** =sct
```{r}
test_output_contains("students2014$kenka", incorrect_msg = "Did you use the $-sign to access the shoe size variable?")

# test if the students code produces an error
test_error()

# Final message the student will see upon completing the exercise
success_msg("Good work!")

```

--- type:NormalExercise lang:r xp:100 skills:1
## Data types (1)

In this section you learn about the data types in R. You will also get familiar with vectors and their usage. What does data types and vectors have to do with each other? A lot, you'll see! 

There are different data types in R. The most commonly used are:

- Numbers and decimal values are called **numeric**
- Text or character values are called **characters**
- [Boolean values](https://en.wikipedia.org/wiki/Boolean_data_type) (`TRUE` or `FALSE`) are called **logical**

When creating character values, one needs to use quotation marks (both `' '` and `" "` will work). For example `"Statistics<3"` would be a character value in R. Numeric and boolean values do not need quotation marks. 

You can use the function `str()` (str is short for structure) to find out what data type the object is.

*** =instructions
- Create character object `my_character` by assigning your initials in it.
- Assign the number 10.5 in `my_numeric_object`. Note that the decimal separator in R is dot instead of comma. 
- Create the object `my_boolean` by executing the line. Note that when creating boolean values, no quotation marks are needed.
- Change the value of `my_boolean` to `TRUE`
- Inspect the data types of the objects with `str()`.

*** =hint
- When creating a character object, remember to use quotation marks.
- Dot is the decimal mark in R:  `2.5` is decimal number but `2,5` means a different thing!
- Remember that `Ctrl + Enter` executes a single row.
- You can use `str()` by putting the object name inside the parenthesis.

*** =pre_exercise_code
```{r}
```

*** =sample_code
```{r}
# Create objects with different data types
my_character <-
my_numeric_object <-
my_boolean <- FALSE

# See how the objects look with the function str()


```

*** =solution
```{r}
# Create objects with different data types
my_character <- "A.B.C."
my_numeric_object <- 10.5
my_boolean <- TRUE

# See how the objects look with the function str()
str(my_character)
str(my_numeric_object)
str(my_boolean)
```

*** =sct
```{r}
test_object("my_character", eval=FALSE, incorrect_msg = "Did you assign your initials to `my_character`?")
test_object("my_numeric_object", incorrect_msg="Did you assign the correct value to `my_numeric_object`?")
test_object("my_boolean", incorrect_msg="Did you change the value of `my_boolean`?")

test_output_contains("str(my_character)", incorrect_msg = "Did you use `str()` on `my_character`?")
test_output_contains("str(my_numeric_object)", incorrect_msg = "Did you use `str()` on `my_numeric_object`?")
test_output_contains("str(my_boolean)", incorrect_msg = "Did you use `str()` on `my_boolean`?")

# test if the students code produces an error
test_error()

# Final message the student will see upon completing the exercise
success_msg("You're making great progress, well done!")
```

--- type:NormalExercise lang:r xp:100 skills:1
## Data types (2)

One of R's data types is called factor. Factors contain numbers, but for every number there is a character value associated to it. A variable measured by the [Likert scale](https://en.wikipedia.org/wiki/Likert_scale) could for example be coded as a factor: the values are numbers from 1 to 5, but the numbers also have words attached to them such as:

- "Strongly disagree" = 1
- "Strongly agree" = 5

In R is is also possible to create factors from numerical objects using the function `cut()`, which splits a variable into categories. It has a mandatory argument `breaks`, which defines the cutting points of the categories.  

*** =instructions
- Create objects `student_age` and `student_age_cut`
- use `str()` to inspect `student_age_cut`
- Use `head()` to print the first six values of `student_age`
- Use `head()` to print the first six values of `student_age_cut`


*** =hint
- Give the object `student_age_cut` as the first argument to `str()` and `head()`.

*** =pre_exercise_code
```{r}
students2014 <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS-data.txt", sep="\t", header=TRUE)

# keep a couple background variables
students2014 <- students2014[,c("sukup","toita","ika","pituus","kenka","kone")]

# recode kone -variable NA values as factor levels
students2014$kone <- addNA(students2014$kone)

# choose rows without missing values
students2014 <- students2014[complete.cases(students2014),]

# integers to numeric
students2014$ika <- as.numeric(students2014$ika)
students2014$pituus <- as.numeric(students2014$pituus)
students2014$kenka <- as.numeric(students2014$kenka)
```

*** =sample_code
```{r}
# students2014 is available

# Let's access the (numerical) age vector of the students2014 and save it to another object.
student_age <- students2014$ika

# Let's use the cut()-function to cut the variable in categories
student_age_cut <- cut(student_age, breaks = c(15, 25, 35, 100))

# Inspect the data type of student_age_cut


# Print the first 6 elements of student_age and student_age_cut



```

*** =solution
```{r}
# students2014 is available

# Let's access the (numerical) age vector of the students2014 and save it to another object.
student_age <- students2014$ika

# Let's use the cut()-function to cut the variable in categories
student_age_cut <- cut(student_age, breaks = c(15, 25, 35, 100))

# Inspect the data type of student_age_cut
str(student_age_cut)

# Print the first 6 elements of student_age and student_age_cut
head(student_age)
head(student_age_cut)


```

*** =sct
```{r}

test_function("str", args="object", incorrect_msg = "Did you use the `str()` to look at the object 'student_age_cut'?")

test_output_contains("head(student_age)", incorrect_msg = "Did you use `head()` and look at the first six elements of `student_age`?")
test_output_contains("head(student_age_cut)", incorrect_msg = "Did you use `head()` to look at the first six elements of `student_age_cut`?")

# test if the students code produces an error
test_error()

# Final message the student will see upon completing the exercise
success_msg("Good work!")

```


--- type:NormalExercise lang:r xp:100 skills:1
## Vectors

A vector is an object containing multiple elements of the same data type. Vectors are quite important in R. The columns of a data frame (such as `student2014`) are vectors.

You can also create vectors by combining values of the same data type, using the function `c()`. 

In this exercise you will learn how to create your own vector.

*** =instructions
- See and execute the examples on how to create a vector with `c()`.
- Create a character vector `my_vector`.
- Create a vector `boolean_vector` with boolean values `TRUE, FALSE, TRUE, FALSE`.
- Print the contents of `boolean_vector`.

*** =hint
- When creating boolean values, no quotation marks are needed.
- You can print the contents of an object by typing it's name.


*** =pre_exercise_code
```{r}
```


*** =sample_code
```{r}
# Here we create a vector with numeric elements
c(2, 3, 4, 5)

# Another one
c(0.1, 0.2, 5.84, 0.7)

# Let's create one with characters and save it to a object
my_vector <- c("a", "b", "c", "d")

# Create boolean vector and print it


```

*** =solution
```{r}
# Here we create a vector with numeric elements
c(2, 3, 4, 5)

# Another one
c(0.1, 0.2, 5.84, 0.7)

# Let's create one with characters and save it to a object
my_vector <- c("a", "b", "c", "d")

# Create boolean vector and print it
boolean_vector <- c(TRUE, FALSE, TRUE, FALSE)
boolean_vector

```

*** =sct
```{r}

test_object("boolean_vector", incorrect_msg = "Did you create `boolean_vector` with same values as instructed?")
test_output_contains("boolean_vector", incorrect_msg = "Did you print the contents of `boolean_vector`?")
test_error()

# Final message the student will see upon completing the exercise
success_msg("Great work!")

```


--- type:NormalExercise lang:r xp:100 skills:1

## Logical comparison

Logical conditions are generally important in programming tasks. Logical comparison creates logical vectors, which can for example be used to select subsets of data. 

The logical comparison operators in R are:  

operator  | description
----------|------------
`==`      | exactly equal to
`!=`      | not equal to
`<`       | less than
`>`       | greater than
`<=`      | less or equal to
`>=`      | greater or equal to

Follow the instructions below to complete the exercise. Take your time, this might not be a quick exercise.

*** =instructions
- Execute and study the example codes
- Add to the row with `c("a","b","c")`: select a suitable logical comparison operator and a suitable comparison value to produce a result vector `FALSE, FALSE, TRUE`
- Add to the row with `c(1,3,2)`: select a suitable logical comparison operator and a suitable comparison value to produce a result vector `TRUE, FALSE, TRUE`
- Bonus: can you think of other ways of achieving the same results?

*** =hint
- The third example should point you in the right direction
- If you're uncertain, just choose a comparison value and try out a bunch of different operators to figure out what might work.


*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# In the comments below: T = TRUE, F = FALSE
# (R also understands these abbreviations)

# Exactly equal
5 == 5 # TRUE

# Not equal
"cat" != "dog" # TRUE

# Greater or equal 3 times
c(0,1,2) >= 1 # F, T, T

# Use logical comparison to produce a result F, F, T
c("a","b","c")

# Use logical comparison to produce a result T, F, T
c(1, 3, 2)


```

*** =solution
```{r}
# In the comments below: T = TRUE, F = FALSE
# (R also understands these abbreviations)

# Exactly equal
5 == 5 # TRUE

# Not equal
"cat" != "dog" # -> TRUE

# Greater or equal 3 times
c(0,1,2) >= 1 # F,T,T

# Use logical comparison to produce a result F,F,T
c("a","b","c") == "c"

# Use logical comparison to produce a result T,F,T
c(1,3,2) < 3

```

*** =sct
```{r}
# submission correctness tests

# example tests:
test_output_contains("c('a','b','c')=='c'", incorrect_msg = "Please make the necessary adustments to produce the output FALSE FALSE TRUE")

test_output_contains("c(1,3,2)<3", incorrect_msg = "Please make the necessary adustments to produce the output TRUE FALSE TRUE")

# test if the students code produces an error
test_error()

# Final message the student will see upon completing the exercise
success_msg("Well done! Remember Spock's words: Logic is the beginning of wisdom, not the end.")

```

--- type:NormalExercise lang:r xp:100 skills:1

## Logical operators

Logical conditions can be combined with the following logical operators:
  
  operator              | description
----------------------- | -----------
  `!`a                  | NOT a
a `&` b                 | a AND b
a <code>&#124;</code> b | a OR b  
  
Logical operators work like logical conditions: they compare the elements of vectors one at a time and can produce logical vectors.

So, for example `F & T` evaluates to `FALSE` and `F | T` evaluates to `TRUE`. Therefore `c(F, T) & c(T, T)` produces a logical vector `FALSE TRUE`.

It is also possible to use parenthesis to control the evaluation of the operators. See how this works from the example codes.

*** =instructions
- Create and print out four subsets of the learning2014 data matching the following conditions. Use the `subset()` function.  
- (gender is male, "M") AND (deep learning is greater than 4.5)
- (age is 25 OR age is 26) AND (attitude is greater than 3.5)
- (gender is female, "N") AND (strategic learning is greater than 4.5)
- (deep learning is more than 4) AND (points is zero OR points is greater than 30)

*** =hint
- The instructions show the correct use of parenthesis when using the logical operators and conditions
- Note that females are coded with "N", not "F".  
- Note that the logical equals is `==`, not `=`.


*** =pre_exercise_code
```{r}
learning2014 <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/learning2014.txt", sep = "\t", header = TRUE)

```

*** =sample_code
```{r}
# learning2014 is available

# male students who scored very high on deep learning
subset(learning2014, (gender == "M") & (deep > 4.5))

# 25 or 26 old students who scored high on attitude
subset(learning2014, (age == 25 | age == 26) & (attitude > 3.5))

# female students who scored very high on strategic learning


# students who scored high on deep learning for whom points is 0 or greater than 30



```

*** =solution
```{r}
# learning2014 is available

# male students who scored very high on deep learning
subset(learning2014, (gender == "M") & (deep > 4.5))

# 25 or 26 old students who scored high on attitude
subset(learning2014, (age == 25 | age == 26) & (attitude > 3.5))

# female students who scored very high on strategic learning
subset(learning2014, (gender == "N") & (stra > 4.5))

# students who scored high on deep learning for whom points is 0 or greater than 30
subset(learning2014, deep > 4 & (points == 0 | points > 30))


```

*** =sct
```{r}
# submission correctness tests

test_output_contains("subset(learning2014,(gender=='N')&(stra>4.5))", incorrect_msg="Please print out the subset containing the female students with high strategic learning scores")
test_output_contains("subset(learning2014,deep>4&(points==0|points>30))", incorrect_msg="Please print out the subset containing the students who scored high on deep learning and had either no points or very high points")

# test if the students code produces an error
test_error()

# Final message the student will see upon completing the exercise
success_msg("Excellent! You are the operator indeed.")

```


--- type:NormalExercise lang:r xp:100 skills:1

## Indices and brackets

Vectors in R store multiple values of the same data type. The values in vectors have indices: the first value in a vector has a index `1`, the second value `2` and so on. 

You can set or get a single value from a vector by using indices and brackets `[ ]`. Using an index number inside the brackets gives access to a single value from the vector. Using a vector of indices gives access to multiple values (another vector). 

It is also possible to rearrange the values in a vector by using indices. Look at the example code to see how indices work. 

*** =instructions
- Create the `names` vector.
- See how the indices work by executing the example lines.
- Use brackets and indices on `names` to create a new vector `girls` with values "Liisa" and "Anna" (in that order).
- Use brackets and indices on `names` to create a new vector `boys` with values "Pekka" and "Heikki" (in that order).

*** =hint
- Note that space between the vector object and bracets produces an error.
- Index vectors `c(1,2)` and `c(2,1)` do not produce the same outcome. The order of the values is different.

*** =pre_exercise_code
```{r}
```


*** =sample_code
```{r}
# Let's create a vector
names <- c("Matti", "Pekka", "Liisa", "Anna")

# Acess the first value of the vector
names[1]

# Change the first value of the vector
names[1] <- "Heikki"

# Acess the 1. and 3. value of the vector
names[c(1, 3)]

# Use indices and brackets to separate the names vector into two vectors of the specified ordering
girls <-
boys <-


  ```

*** =solution
```{r}
# Let's create a vector
names <- c("Matti", "Pekka", "Liisa", "Anna")

# Acess the first value of the vector
names[1]

# Change the first value of the vector
names[1] <- "Heikki"

# Acess the 1. and 3. value of the vector
names[c(1, 3)]

# Use indices and brackets to separate the names vector into two vectors of the specified ordering
girls <- names[c(3, 4)]
boys <- names[c(2, 1)]

```

*** =sct
```{r}

test_object("girls", incorrect_msg = "Did you create `girls`?")
test_object("boys", incorrect_msg = "Did you create `boys`? Are the values in the correct order?")

test_error()

# Final message the student will see upon completing the exercise
success_msg("Such indices. Wow. Much intelligent.")

```


--- type:NormalExercise lang:r xp:100 skills:1

## Easy vectors

Sometimes you need a very long index vector. In that case it would be a lot of work to create the vector by combining integer values with `c()`. Luckily, there are convenient ways to create long vectors in R.

For integer values, a specially useful one is the method `start:end`, which creates an integer vector with all the values from start to end. The following two lines hence produce identical results

```
1:5  
c(1,2,3,4,5)
```


*** =instructions
- Create an integer vector with values from 1 to 5.
- Create an integer vector with even values from 2 to 10.
- Create object `attitude` and give the values in it names matching the indices.
- Access index values 1-5 of `attitude`
- Use `:` to create an integer vector with the values 1, 2, ..., 10
- Access every second value of the `attitude` vector, starting from the 2. value until the 20th value. These values correspond to the even numbered indeces of the vector: 2, 4, .. , 20

*** =hint
- First you will need an index vector with values 2, 4, .. , 20. The example shows how to create such a vector
- You can then use the index vector together with brackets (`[ ]`) to complete the task


*** =pre_exercise_code
```{r}
learning2014 <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/learning2014.txt", sep = "\t", header = TRUE)
```

*** =sample_code
```{r}
# learning2014 is available

# Create an integer vector with values 1,2,..,5
1:5

# Create an integer vector with even values 2, 4, .. , 10
(1:5)*2

# Create object attitude and give the data points names 1, 2, ..
attitude <- learning2014$attitude
names(attitude) <- 1:length(attitude)

# Access the values 5 - 10 of attitude
attitude[5:10]

# Create an integer vector with values 1,2,..,10


# Access every second value of attitude from 2. to the 20th index


```

*** =solution
```{r}
# learning2014 is available

# Create an integer vector with values 1,2,..,5
1:5

# Create an integer vector with even values
(1:5)*2

# Create object attitude and give the data points names 1, 2, ..
attitude <- learning2014$attitude
names(attitude) <- 1:length(attitude)

# Access the values 5 - 10 of attitude
attitude[5:10]

# Create an integer vector with values 1,2,..,10
1:10

# Access every second value of attitude from 2. to the 20th index
attitude[(1:10)*2]


```

*** =sct
```{r}
# submission correctness tests

test_student_typed("1:10", not_typed_msg = "Did you use `:` to create and print out the specified integer vector?")
test_output_contains("attitude[(1:10)*2]")

# test if the students code produces an error
test_error()

# Final message the student will see upon completing the exercise
success_msg("Great work!")

```

